Bootstrap learning via modular concept discovery

### 2013-02-01 drafts/ijcai2013/ijcai13.tex
Changes.

@@

\author{Eyal Dechter \\
MIT\\
USA \\
edechter@mit.edu
\And
Jon Malmaud \\
MIT \\ 
USA \\
malmaud@mit.edu
\And 
Ryan P.  Adams \\
Harvard University\\
USA \\
rpa@seas.harvard.edu
\And
Joshua B. Tenenbaum \\
MIT\\
USA \\
jbt@mit.edu
}

@@

\maketitle

\begin{abstract}
 Suppose a learner is faced with a domain of problems about which it
 knows nearly nothing. It does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer it effectively. How can such a learner ever get off
 the ground? A common intuition is that if the solutions to these
 problems share a common structure, and the learner can solve some
 simple problems by brute force, it should be able to extract useful
 components from these solutions and, by composing them, explore the
 solution space more efficiently. Here, we formalize this intuition,
 where the solution space is the space of typed functional programs
 and the gained information is stored as a stochastic grammar over
 programs. We propose an iterative procedure for exploring such
 spaces: in the first step of each iteration, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. We test this
 procedure on symbolic regression and boolean circuit learning and
 show that the learner discovers modular concepts for these
 domains. Whereas the learner is able to solve almost none of the
 posed problems in the procedure's first iteration, it rapidly becomes
 able to solve a large number by gaining abstract knowledge of the
 structure of the solution space.

@@

Recall that a valid partial solution $\mathcal{H}$ is a subtree of
$\mathcal{G}$ satisfying the following properties: the root of
$\mathcal{H}$ is the root of $\mathcal{G}$ and any OR node in
$\mathcal{H}$ has at most one child. $\mathcal{H}$ is a complete
solution (and thus a complete expression) if it is a partial solution
whose leaves are leaves of $\mathcal{G}$, if each OR node in
$\mathcal{H}$ has exactly one child, and if each AND node in
$\mathcal{H}$ is connected to all of that node's children in
$\mathcal{G}$

The value of a partial solution $\mathcal{H}$ is calculated by summing
the value of its leaves, where each such value is the log of the
probability assigned to that leaf in
Section~\ref{sec:stochgrammar}. If a leaf of $\mathcal{H}$ is a leaf
of $\mathcal{G}$ then either it is an AND node, to which we assign a
value of 0, or it corresponds to some primitive combinator $c_i$ in
the library and is the child of some typed OR node with type $t$. We
assign it a value of $\log{p(c_i | t)}$. If a leaf $v$ of
$\mathcal{H}$ is an OR node, then the value of any extension of it to
a complete solution is at most $log(1/4)$. Thus, we have defined a
value for partial solutions that upper bounds the value of any of its
extensions.

Our best-first policy is to expand the next partial solution
with the highest value. If a partial solution has multiple leaves that
are not primitive combinators, we expand the left most one first,
since it is the choice of left combinator that defines the type of its
right combinator in an application pair. That is, the parent node
applies its constraint on valid solutions via the left child, not the
right child. We want to apply constraints as early as possible, so
expanding left children first prunes more partial solutions.

@@

A natural metric for the compressibility of a set of binary trees is
the number of unique subtrees (counting leaves as subtrees) that this
set contains. As an example of this, suppose we had six tasks whose
solutions were integers $4^2, 5^2, 6^2, 7^2, 8^2$ and suppose our
frontier contained the 10 expressions:

\captionsetup[subfigure]{labelformat=empty} 
\begin{figure}[h!]
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7, background rectangle/.style={fill=lightgray}] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 4 ] 4 ]
     \end{tikzpicture}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 5 ] 5 ]
     \end{tikzpicture}
  \end{subfigure}
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture} [scale=.7]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 6 ] 6 ]
     \end{tikzpicture}
  \end{subfigure}
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 7 ] 7 ]
     \end{tikzpicture}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 8 ] 8 ]
     \end{tikzpicture}
  \end{subfigure}
  \vspace{2mm}

@@

  \vspace{2mm}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 4 ]
     \end{tikzpicture}
     \caption{(16)}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 5 ]
     \end{tikzpicture}
     \caption{(25)}
  \end{subfigure}
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture} [scale=.6]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 6 ]
     \end{tikzpicture}
     \caption{(36)}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 7 ]
     \end{tikzpicture}
     \caption{(49)}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 8 ]
     \end{tikzpicture}
     \caption{(64)}
  \end{subfigure}
\end{figure}
\captionsetup[subfigure]{labelformat=parens}

The top row consists of one type of representation of an integer
squared, namely, the integer times itself. The bottom row contains
another type of representation in which the function $f(x)=x^2$,
represented in combinatory logic as $S * I$, is applied to the
integer. The top and bottom row of each expression evaluates to the
same integer. If we consider only the first column, the top tree has 4
unique subtrees and the bottom has 7. But each column adds 3 unique
subtrees to the top row and only 2 to the bottom. So by the fifth
column the top row has more unique subtrees than the bottom row. The
increased compressibility of the bottom row is due to the fact that
the squaring operation has been factored out as an abstraction. This
is an example of how this compressibility metric favors
abstraction when multiple expressions are being considered at once.

We want to choose a solution for each hit task such that the complete set
of such solutions for all solved tasks is as compressible as
possible. That is, let $t_i$ be a task in the set of hit tasks, and
let $\{e^k_i\}$ be the expressions in the frontier that solves
it. Then our goal is to find
\[
\{e^*_i\} = \argmin_{\{e^k_i\}} | \cup_i e^k_i |,
\]
where $|\cdot|$ denotes the number of unique trees in a set of expressions.

However, an exact optimization of this problem requires examining
$O(M^K)$ assignments, where $M$ is the maximum number of solutions for
any one task and $K$ is the number of tasks. Since this is
prohibitive, we relax the compressibility metric as follows: we define
the cost of adding each new solution $s_n$ to a partial set of
solutions $s_1, \dots, s_{n-1}$ to be the number of additional unique
subtrees in $s_n$ that are not present in $s_{n-1}$. That is, the
penalty of adding a new expression to the set of expressions depends
only on the last expression we added. Our solution thus becomes
approximate and that approximation is order-dependent, but a
depth-first search on the defined search space goes from exponential
in the solution degree to quadratic ($O(KM^2)$).

@@

Given a set of solution programs, we want to re-estimate our
stochastic grammar. Our inspiration for doing this is the
Neville-Manning algorithm for grammar-based
compression~\cite{nevill1997identifying}. The idea of that compression
algorithm is to compress any string by creating a grammar for that
string such that a) no two symbols appear more than once in the
grammar and b) every rule is used more than once. From a set of trees
we generate a grammar according to these criteria.

@@

This procedure generates a parse of the solution set with a new set of
primitives, and we keep count of how many times each primitive
occurred in the solutions set. To estimate the probabilities
associated with each node production, we again traverse the solution
set, but this time for each node we count which other primitive
elements from the new grammar could have been used. In accordance with
Section~\ref{stochgrammar}, we estimate the terminal production weight
to be the ratio of the number of times that node was used in the first
traversal divided by the number of times it was used in the second
traversal.

@@

We first illustrate the performance of our learning algorithm on a
symbolic regression problem. In task in this problem, we want to find
an algebraic function, symbolically specified, that maps a set of
input values to output values. We choose this problem because it has a
long history in AI, particularly in the genetic programming
literature~\cite{DBLP:books/daglib/0070933}.

In the formulation we consider, each task $t$ corresponds to a
polynomial $f$, and an expression $e$ solves $t$ if ${ (e\; i)=f(i) }$
for all ${ i=0,\dots,9 }$. For this experiment, the set of problems
corresponds to the set of all polynomials with degree less than or
equal to 2. In the initial library, we include the basic combinators
I, S, B, C and four arithmetic primitives, $1$, $0$, $*$, and $+$. The
choice of initial weights on these primitives can make a difference,
as it determines the relative ordering of combinators in the
enumeration step. To get a general sense of the algorithm's
performance, we set the initial weights to be slightly different on
each run, perturbing them around a uniform weighting.

\setcounter{subfigure}{0} 
\begin{figure}[ht]
\centering
\begin{subfigure}[Before]{\linewidth}
\includegraphics[width=.9\linewidth]{figures/symreg3frontiers.pdf}
\caption{Learning curves as a function of frontier size.  As frontier
  size is increased, it plateaus closer to $100\%$ performance. A
  control brute-force search over $150000$ only hits $3\%$ of the
  tasks. \label{fig:symregLearning}}
\end{subfigure}

@@

In Figure~\ref{fig:symregLearning}, we show performance results as we vary the
frontier size. Each individual plot shows how changing the frontier
size changes the number of tasks the algorithm identifies correctly
over the course of fifteen algorithm iterations. As a control, we
enumerated $10000*15=150000$ expressions from the initial
grammar. This is the total number of expressions that a run of the
algorithm sees if it has a frontier size of $10000$ and runs for $15$
iterations. Thus, if the benefit accruing to the runs with larger
frontier sizes is due to simple an increase in brute-force search
ability, rather than increased learning, we should see similarly good
performance from this control run. In fact, however, the control run
only hit $27$ of the $1000$ asks ($3\%$), whereas our algorithm nears $100\%$.

What kinds of primitives does the E-C algorithm learn in this task?
Inspecting the top weighted primitives of the final grammars, we find
many incrementers (e.g. (+1), (+2), etc.), several versions of
expressions that translate to functions like $x*f(x)$ and $f(x*f(x))$,
and combinations of these, like $x*(x+3)+3$ and $x*(x+2)$. Informally,
we see expressions that apply functions to themselves in various ways
several times, building up complex functions with relatively few
unique primitives. In the next experiment we will inspect more
directly the modular components that are learned by this algorithm.

We would like to assess to what degree ``conceptual bootstrapping'' is
responsible for the improved performance, that is, to what extent does
the presence of simple problems account for ability to learn complex
functions.  One prediction of this idea is that to succeed on a set of
tasks, the set must contain a ``learnable'' curriculum, a set of tasks
that serve as stepping stones from an impoverished representation to a
rich one. We can test this prediction by varying the set of tasks to
which the E-C algorithm is exposed.

In Figure~\ref{fig:symregCurricula}, we present learning curves
(frontier size 10000) corresponding to various versions of our
original symbolic regression task set. Recall that the original set
consisted of all polynomials of degree two or less and with
coefficients between 0 and 9. In the ``no constant'' task set, we
remove the 9 constant functions in this set. In the ``no linear'' task
set, we remove 90 linear functions from the set. We observe that
performance on those task sets does not decrease. However, when we
remove both the linear functions and the constant function (``only
quadratics''), we see a sharp drop in the algorithm's
performance. When we further restrict the task set to only ``complex''
quadratics (which we are defining as quadratics all of whose
coefficients are greater than zero), we observe another comparable
drop in performance. When we go one step further and restrict the task
set to quadratics with coefficients greater than 1, performance drops
to 0 because no task is hit in the initial frontier.

@@

\subsection{Boolean circuits}
\label{sec:boolcircuits}

\begin{figure}[ht]
\includegraphics{./tablealonedoc.pdf}
\vspace{-.5cm}
\caption{Some learned primitives in Boolean function learning
  experiment. We highlight compression using $NOT$, $E_1$ and $E_2$ in
  some of the expressions they are found. \,(a)The three elementary
  gates used to generate the task set. (b) Two higher-order primitives
  in the induced grammer. Note how $E_1$ can be used to define FALSE
  by taking the NOT function as an
  argument.  \label{table:booleanCircuits}}
\vspace{-.4cm}
\end{figure}

As another demonstration of this approach, we investigate the
procedure's ability to learn boolean functions using only the logical
primitive NAND and the basic combinators. It is well known that a
boolean circuit can be constructed for any boolean function given only
the NAND gate. Here, the basic combinators take the place of the
wiring normally found in a circuit, the output of element to multiple
inputs downstream.

@@

We constructed two task sets. The first of these was constructed
explicitly to contain familiar module structure. In this set of tasks,
we sampled $1000$ boolean circuits using AND, OR, and NOT gates. To
accomplish this, we first sampled either 1, 2, 3 inputs, then between
1 and 5 gates, randomly wiring the inputs of each new gate to the
output of one of the existing gates in the circuit. We continued this
sampling procedure until we had 1000 ``connected'' circuits, i.e.,
circuits all of whose output are wired to an input except for the last
one (the output gate). This resulted in $1000$ tasks consisting of
$82$ unique boolean functions, with a distribution of tasks as shown
in Figure~\ref{fig:booldistr}. In Figure~\ref{fig:boolLearningCurvesCircuits},
We present learning curves for frontier sizes 100, 500 and 10000,
using the same procedure as in Section~\ref{sec:symreg}.

\begin{figure}
\vspace{-.3cm}
\includegraphics[width=\linewidth]{figures/circuitDistrComp.pdf}
\vspace{-.6cm}
\caption{Distribution of boolean functions among (a) the set of 1000
  circuits randomly constructed from AND, OR and NOT gates (see text);
  b) the first 1000 expressions from the grammar over programs
  \emph{before} learning; c) the first 1000 expressions from the
  grammar over programs \emph{after} learning. Gray levels indicate
  percent of instances (circuits/expressions) which evaluate to that
  function.\label{fig:booldistr}}
\end{figure}

This problem is particularly suitable for inspecting the constituents
of the induced grammar $\mathcal{G}$. We might hope that our algorithm
recovers the constituent logic gates that were used to build up the
problems. Table~\ref{table:bool} shows a few of the top ranked
expressions in the library. These include the basic logic gates; we
also include one higher-order expression $F$, to stress that the
representation we are using allows one to build concepts that are more
expressive that just sub-circuits.

In a second experiment, we include all $272$ boolean truth tables with
three or fewer inputs as tasks. This is a more difficult problem. Note
that in the first boolean function problem, there are two kinds of
learning that can take place: first, many expressions map to a single
boolean function, so we need to learn primitives that allow us to span
many different functions instead of allows generating very simple
functions over and over again. The second kind of learning involves
the distribution of the functions themselves. In the first boolean
experiment, that structure is apparent in the distribution in
Figure~\ref{fig:boolhist}. In this second experiment, we remove the
second kind of structure. In Figure~\ref{fig:boolLearningCurvesUniform}, we show
10 runs of the algorithm with a frontier size of $2000$: note how
there are several local minima that majority of the runs get stuck in
with performance around $50\%$, but several of the runs seem to take
different trajectories to more successful representations.

@@

\vspace{-.2cm}
\begin{subfigure}{.1\linewidth}
\caption{\label{fig:boolLearningCurvesUnifrom}}
\end{subfigure}
\begin{subfigure}[After]{.9\linewidth}
\includegraphics[width=.9\linewidth]{figures/booleanfunctionsUniform.pdf}

\end{subfigure}
\vspace{-.3cm}
\caption{Learning curves from Boolean function learning
  experiment. (a) Learning curves for various frontier sizes on task
  set of Boolean functions drawn randomly from a distribution (see
  text for details) of circuits built up of NOT, AND, OR gates. (b) 10
  learning curves for frontier size of 2000 on a task set consisting
  of all Boolean functions of cardinality 3 or smaller. Note how most
  of the trajectories get stuck in a local minimum around
  $50\%$.\label{fig:boolLearningCurves}

@@

%% \subsection{Digital arithmetic}
%% Having demonstrated the behavior of the E-C algorithm on symbolic
%% regression, we analyze its performance on tasks involving digital
%% arithmetic, a class of problems with known modular structure. The
%% general goal is to discover boolean circuits consisting of a limited
%% set of gates that perform arithmetic operations on binary number
%% representations.

\section{Discussion}
This work attempts to bring together ideas from various areas of
research -- hierarchical Bayes, learning to learn, minimum description
length learning -- to provide a proof of concept that learning
abstractions from programs via compression can rapidly improve search
performance. We emphasize this by focusing on domains in which the
solution space is not smooth and the error signal is all or
none. Thus, the main source of information for learning is the modular
and compositional structure of the solution space.

>>>>

\usepackage{titling}

\setlength{\droptitle}{0em}   
\date{}

@@

%% \author{Eyal Dechter \\
%% MIT\\
%% USA \\
%% edechter@mit.edu
%% \And
%% Jon Malmaud \\
%% MIT \\ 
%% USA \\
%% malmaud@mit.edu
%% \And 
%% Ryan P.  Adams \\
%% Harvard University\\
%% USA \\
%% rpa@seas.harvard.edu
%% \And
%% Joshua B. Tenenbaum \\
%% MIT\\
%% USA \\
%% jbt@mit.edu
%% }

@@

\maketitle
\begin{abstract}
 Suppose a learner is faced with a domain of problems about which it
 knows nearly nothing. It does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer it effectively. How can such a learner ever get off
 the ground? A common intuition is that if the solutions to these
 problems share a common structure, and the learner can solve some
 simple problems by brute force, it should be able to extract useful
 components from these solutions and, by composing them, explore the
 solution space more efficiently. Here, we formalize this intuition,
 where the solution space is the space of typed functional programs
 and the gained information is stored as a stochastic grammar over
 programs. We propose an iterative procedure for exploring such
 spaces: in the first step of each iteration, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. We test this
 procedure on symbolic regression and boolean circuit learning and
 show that the learner discovers modular concepts for these
 domains. Whereas the learner is able to solve almost none of the
 posed problems in the procedure's first iteration, it rapidly becomes
 able to solve a large number by gaining abstract knowledge of the
 structure of the solution space.

@@

Recall that a valid partial solution $\mathcal{H}$ is a subtree of
$\mathcal{G}$ satisfying the following properties: the root of
$\mathcal{H}$ is the root of $\mathcal{G}$ and any OR node in
$\mathcal{H}$ has at most one child. $\mathcal{H}$ is a complete
solution (and thus a complete expression) if it is a partial solution
whose leaves are leaves of $\mathcal{G}$, if each OR node in
$\mathcal{H}$ has exactly one child, and if each AND node in
$\mathcal{H}$ is connected to all of that node's children in
$\mathcal{G}$.

The value of a partial solution $\mathcal{H}$ is calculated by summing
the value of its leaves, where each such value is the log of the
probability assigned to that leaf in
Section~\ref{sec:stochgrammar}. If a leaf of $\mathcal{H}$ is a leaf
of $\mathcal{G}$ then either it is an AND node, to which we assign a
value of 0, or it corresponds to some primitive combinator $c_n$ in
the library and is the child of some typed OR node with type $\tau$. We
assign it a value of $\log{p(c_n | \tau)}$. If a leaf $v$ of
$\mathcal{H}$ is an OR node, then the value of any extension of it to
a complete solution is at most $log(1/4)$. Thus, we have defined a
value for partial solutions that upper bounds the value of any of its
extensions.

Our best-first policy is to expand the next partial solution
with the highest value. If a partial solution has multiple leaves that
are not primitive combinators, we expand the left most one first,
since it is the choice of left combinator that defines the type of its
right combinator in an application pair. That is, the parent node
applies its constraint on valid solutions via the left child, not the
right child. We want to apply constraints as early as possible, so
expanding left child first prunes more partial solutions.

@@

A natural metric for the compressibility of a set of binary trees is
the number of unique subtrees (counting leaves) that this
set contains. As an example of this, suppose we have six tasks whose
solutions are integers $4^2, 5^2, 6^2, 7^2$, and $8^2$, and suppose our
frontier contains the $10$ expressions:
\vspace{-.3cm}
\captionsetup[subfigure]{labelformat=empty} 
\begin{figure}[h!]
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7, background rectangle/.style={fill=lightgray}] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 4 ] 4 ]
     \end{tikzpicture}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 5 ] 5 ]
     \end{tikzpicture}
  \end{subfigure}
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture} [scale=.7]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 6 ] 6 ]
     \end{tikzpicture}
  \end{subfigure}
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 7 ] 7 ]
     \end{tikzpicture}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.7] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[  * 8 ] 8 ]
     \end{tikzpicture}
  \end{subfigure}
  \vspace{2mm}

@@

  \vspace{2mm}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 4 ]
     \end{tikzpicture}
     \caption{(16)}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 5 ]
     \end{tikzpicture}
     \caption{(25)}
  \end{subfigure}
   \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture} [scale=.6]
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 6 ]
     \end{tikzpicture}
     \caption{(36)}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 7 ]
     \end{tikzpicture}
     \caption{(49)}
  \end{subfigure}
  \begin{subfigure}{.17\linewidth}
     \begin{tikzpicture}[scale=.6] 
       \tikzset{level distance=15pt, sibling distance=10pt}
       \Tree [[[  S * ] I ] 8 ]
     \end{tikzpicture}
     \caption{(64)}
  \end{subfigure}
\end{figure}
\captionsetup[subfigure]{labelformat=parens} 
\vspace{-.3cm}
The top row consists of one type of representation of an integer
squared, namely, the integer times itself. The bottom row contains
another type of representation in which the function $f(x)=x^2$,
represented in combinatory logic as $S * I$, is applied to the
integer. The top and bottom row of each expression evaluate to the
same integer. If we consider only the first column, the top tree has 4
unique subtrees and the bottom has 7. But each column adds 3 unique
subtrees to the top row and only 2 to the bottom. So by the fifth
column the top row has more unique subtrees than the bottom row. The
increased compressibility of the bottom row is due to the fact that
the squaring operation has been factored out as an abstraction. This
is an example of how this compressibility metric favors
abstraction when multiple expressions are being considered at once.

We want to choose a solution for each hit task such that the complete set
of such solutions for all solved tasks is as compressible as
possible. That is, let $t_i$ be a task in the set of hit tasks, and
let $\{e^k_i\}$ be the expressions in the frontier that solve
it. Then our goal is to find
\[
\{e^*_i\} = \argmin_{\{e^k_i\}} | \cup_i e^k_i |,
\]
where $|\cdot|$ denotes the number of unique trees in a set of expressions.

However, an exact optimization of this problem requires examining
$O(M^K)$ assignments, where $M$ is the maximum number of solutions for
any one task, and $K$ is the number of tasks. Since this is
prohibitive, we relax the compressibility metric as follows: we define
the cost of adding each new solution $s_n$ to a partial set of
solutions $s_1, \dots, s_{n-1}$ to be the number of additional unique
subtrees in $s_n$ that are not present in $s_{n-1}$. That is, the
penalty of adding a new expression to the set of expressions depends
only on the last expression we added. Our solution thus becomes
approximate, and that approximation is order-dependent, but a
depth-first search on the defined search space goes from exponential
in the solution degree to quadratic ($O(KM^2)$).

@@

Given a set of chosen solution expressions, which we will call the
\emph{solution set}, want to re-estimate our stochastic grammar. Our
inspiration for doing this is the Neville-Manning algorithm for
grammar-based compression~\cite{nevill1997identifying}. The idea of
that compression algorithm is to compress any string by creating a
grammar for that string such that a) no two symbols appear more than
once in the grammar and b) every rule is used more than once. From a
set of expressions we generate a grammar according to these criteria.

@@

This procedure generates a parse of the solution set with a new set of
primitives, and we keep count of how many times each primitive
occurred in the solutions set. To estimate the probabilities
associated with each node production, we again traverse the solution
set, but this time for each node we count which other primitive
elements from the new grammar could have been used. In accordance with
Section~\ref{sec:stochgrammar}, we estimate the terminal production weight
to be the ratio of the number of times that node was used in the first
traversal divided by the number of times it was used in the second
traversal.

@@

We first illustrate the performance of our learning algorithm on a
symbolic regression problem. For each task in this problem, we want to
find an algebraic function, symbolically specified, that maps a set of
input values to output values. We choose this problem because it has a
long history in AI, particularly in the genetic programming
literature~\cite{DBLP:books/daglib/0070933}.

In the formulation we consider, each task $t$ corresponds to a
polynomial $f$, and an expression $e$ solves $t$ if it returns the
same value when applied $i \in 0, \dots, 9$ that $f$ does. In the
experiment shown here, the set of problems corresponds to the set of
all polynomials with degree less than or equal to 2. In the initial
library, we include the basic combinators I, S, B, C and four
arithmetic primitives, $1$, $0$, $*$, and $+$. The choice of initial
weights on these primitives can make a difference, as it determines
the relative ordering of combinators in the enumeration step. To get a
general sense of the algorithm's performance, we set the initial
weights to be slightly different on each run, perturbing them around a
uniform weighting.

\setcounter{subfigure}{0} 
\begin{figure}[ht]
\centering
\begin{subfigure}[Before]{\linewidth}
\includegraphics[width=.9\linewidth]{figures/symreg3frontiers.pdf}
\caption{Learning curves as a function of frontier size.  As frontier
  size is increased, curves plateau closer to $100\%$ performance. A
  baseline search over $150000$ expressions only hits $3\%$ of the
  tasks. \label{fig:symregLearning}}
\end{subfigure}

@@

In Figure~\ref{fig:symregLearning}, we show performance results as we
vary the frontier size.Changing the frontier size changes the number
of tasks the algorithm identifies correctly over the course of 15
algorithm iterations (all experiments in this work were run for 15
iterations). As a baseline, we enumerated $10000*15=150000$
expressions from the initial grammar. This is the total number of
expressions that a run of the algorithm sees if it has a frontier size
of $10000$ and runs for $15$ iterations. Thus, if the benefit accruing
to the runs with larger frontier sizes is due to simply an increase in
the number of expressions seen, rather than increased learning, we should
see similarly good performance from this baseline run. In fact,
however, the baseline run only hits $27$ of the $1000$ tasks ($3\%$),
whereas our algorithm nears $100\%$ for a frontier of $10000$.

What does the E-C algorithm learn in this task?  Inspecting the top
weighted primitives of the final grammars, we find many incrementers
(e.g. (+1), (+2), etc.), several versions of expressions that
translate to functions like $x*f(x)$ and $f(x*f(x))$, and combinations
of these, like $x*(x+3)+3$ and $x*(x+2)$. Informally, we see
expressions that apply functions to themselves, building up complex
functions with relatively few unique primitives.

To what degree is ``conceptual bootstrapping'' responsible for the
improved performance. That is, to what extent does the presence of
simple problems account for the ability to learn complex functions?
One hypothesis is that to succeed on a set of tasks, the set must
contain a ``learnable'' curriculum, a set of tasks that serve as
stepping stones from an impoverished representation to a rich one. We
can test this hypothesis by varying the set of tasks to which the EC
algorithm is exposed. If it is true, then we should see a nonlinear
reponse to reducing the number of easy tasks, as the curriculum
changes from being learnable to being unlearnable.

In Figure~\ref{fig:symregCurricula}, we present learning curves
(frontier size 10000) corresponding to various versions of our
original symbolic regression task set. Recall that the original set
consisted of all polynomials of degree two or less and with
coefficients between 0 and 9. In the ``no constant'' task set, we
remove the 9 constant functions in this set. In the ``no linear'' task
set, we remove 90 linear functions from the set. We observe that
performance on those task sets does not decrease. However, when we
remove both the linear functions and the constant function (``only
quadratics''), we see a sharp drop in the algorithm's
performance. When we further restrict the task set to only ``complex''
quadratics (which we define as quadratics whose
coefficients are greater than zero), we observe another comparable
drop in performance. When we go one step further and restrict the task
set to quadratics with coefficients greater than 1, performance drops
to 0 because no task is hit in the initial frontier.

@@

\subsection{Boolean function learning}
\label{sec:boolcircuits}

\begin{figure}[ht]
\includegraphics{./tablealonedoc.pdf}
\vspace{-.5cm}
\caption{Some learned primitives from the circuit based Boolean
  function learning experiment. We highlight compression using $NOT$,
  $E_1$ and $E_2$. \,(a)The three elementary gates used to generate
  the task set. (b) Two higher-order primitives in the induced
  grammer. Note how $E_1$ can be used to define FALSE by taking the
  NOT function as an argument. (c) Additional common Boolean functions
  found in the learned grammar.  \label{table:booleanCircuits}}
\vspace{-.4cm}
\end{figure}

As another demonstration of our approach, we investigate the
EC algorithm's ability to learn Boolean functions using only the logical
primitive NAND and the basic combinators. It is well known that a
boolean circuit can be constructed for any boolean function given only
the NAND gate. Here, the basic combinators take the place of the
wiring normally found in a circuit.

@@

We constructed two task sets. The first of these was constructed
explicitly to contain familiar modular structure. In this set of
tasks, we sampled $1000$ Boolean circuits using AND, OR, and NOT
gates. To accomplish this, we first sampled either 1, 2, 3 inputs,
then between 1 and 5 gates, randomly wiring the inputs of each new
gate to the output of one of the existing gates in the circuit. We
continued this sampling procedure until we had 1000 ``connected''
circuits, i.e., circuits all of whose outputs are wired to an input
except for the last one (the output gate). This resulted in $1000$
tasks consisting of $82$ unique Boolean functions, with a distribution
of tasks as shown in Figure~\ref{fig:booldistr}~(a). In
Figure~\ref{fig:boolLearningCurvesCircuits}, We present learning
curves for frontier sizes 100, 500 and 10000, using the same procedure
as in Section~\ref{sec:symreg}. In Figure~\ref{fig:booldistr}, we show
that the distribution over Boolean functions enumerated from the
grammar over expressions is much more similar to the true function
distribution after learning than before.

\begin{figure}
\vspace{-.3cm}
\includegraphics[width=\linewidth]{figures/circuitDistrComp.pdf}
\vspace{-.6cm}
\caption{Distribution of boolean functions among (a) the set of 1000
  circuits randomly constructed from AND, OR and NOT gates (see text);
  b) the first 1000 expressions from the grammar over expressions
  \emph{before} learning; c) the first 1000 expressions from the
  grammar over expressions \emph{after} learning. Gray levels indicate
  percent of instances (circuits/expressions) which evaluate to that
  function.\label{fig:booldistr}}
\end{figure}

This problem is particularly suitable for inspecting the constituents
of the induced grammar $\mathcal{G}$. We might hope that our algorithm
recovers the constituent logic gates that were used to build up the
tasks. Table~\ref{table:booleanCircuits} shows a few of the top ranked
expressions in the library. These include the basic logic gates; we
also include two higher-order expression $E_1$ and $E_2$, to stress
that the representation the EC algorithm is using allows it to build
concepts that are more expressive that just sub-circuits.

In a second experiment, we include all $272$ Boolean truth tables with
three or fewer inputs as tasks. This is a more difficult
problem. There are two kinds of learning that the EC algorithm might
accomplish: first, many expressions in $\mathcal{L}$ map to a single
Boolean function, so it needs to learn primitives that allow it to
span many different functions instead of generating very simple
functions redundantly. The second kind of learning involves the
distribution of the functions themselves. In the circuit based Boolean
experiment, that structure is apparent in the distribution in
Figure~\ref{fig:booldistr}. In this second experiment, we remove the
second kind of structure. In
Figure~\ref{fig:boolLearningCurvesUniform}, we show 10 runs of the EC
algorithm on the second experiment with a frontier size of $2000$: note how there are several
local minima that the majority of the runs get stuck in with performance
around $50\%$, but several of the runs seem to take different
trajectories to more successful representations.

@@

\vspace{-.2cm}
\begin{subfigure}{.1\linewidth}
\caption{\label{fig:boolLearningCurvesUniform}}
\end{subfigure}
\begin{subfigure}[After]{.9\linewidth}
\includegraphics[width=.9\linewidth]{figures/booleanfunctionsUniform.pdf}

\end{subfigure}
\vspace{-.3cm}
\caption{Learning curves from Boolean function learning
  experiments. (a) Learning curves for various frontier sizes on task
  set of circuit based Boolean functions. (b) 10
  learning curves for frontier size of 2000 on a task set consisting
  of all Boolean functions of cardinality 3 or smaller. Note how most
  of the trajectories get stuck in a local minimum around
  $50\%$.\label{fig:boolLearningCurves}

@@

\section{Conclusion}
This work brings together ideas from various areas of research --
hierarchical Bayes, learning to learn, minimum description length
learning, and learning programs -- to provide a proof of concept for
how bootstrap learning of abstract composable concepts can rapidly
improve search performance.  We show that this approach can learn
successfully even in domains for which the solution space is not
smooth and the error signal is all or none, where the only sense of
locality to guide learning is the modular and compositional structure
of the solutions that the algorithm itself builds.
